% Created 2019-04-17 Wed 12:42
% Intended LaTeX compiler: pdflatex
\documentclass[a4paper,12pt,twoside,openany]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage[acronym]{glossaries}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage{datetime2}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{float}
\usepackage{amssymb}
\usepackage{verbatim}
\usepackage{caption}
\usepackage[linesnumbered,ruled]{algorithm2e}
\usepackage{listings}
\usepackage{minted}
\usepackage[pdfborder={0 0 0}]{hyperref}    % turns references into hyperlinks
\usepackage{docmute}   % only needed to allow inclusion of proposal.tex
\usepackage[margin=25mm]{geometry}
\date{\today}
\title{Implementing Parallelism in Lisp for REDUCE}
\author{Andrei-Vlad Badelita}

\raggedbottom                           % try to avoid widows and orphans
\sloppy
\clubpenalty1000%
\widowpenalty1000%

\renewcommand{\baselinestretch}{1.1}    % adjust line spacing to make
% more readable

\DTMusemodule{british}{en-GB}
\DTMsetregional
\DTMlangsetup{showdayofmonth=false}

\graphicspath{ {figures/} }

\usepackage[backend=biber,sorting=none]{biblatex}
\addbibresource{references.bib}

\newenvironment{code}{\captionsetup{type=listing}}{}

% \makeglossaries

\begin{document}

% Title
%==========================================================

\hfill{\LARGE \bf Andrei-Vlad Bădeliță}

\pagestyle{empty}

\vspace*{60mm}
\begin{center}
\Huge{\bf Implementing Parallelism in Lisp for REDUCE} \\
\vspace*{5mm}
Part II Computer Science Dissertation \\
\vspace*{5mm}
Trinity College \\
\vspace*{5mm}
\today  % today's date
\end{center}

\cleardoublepage

%==============================================

\pagestyle{plain}
\pagenumbering{roman}

\section*{Declaration of originality}

I, Andrei-Vlad Bădeliță of Trinity College, being a candidate for Part II of the
Computer Science Tripos, hereby declare that this dissertation and the work described
in it are my own work, unaided except as may be specified below, and that the
dissertation does not contain material that has already been used to any
substantial extent for a comparable purpose.

\medskip
\noindent
I, Andrei-Vlad Bădeliță of Trinity College, am content for my dissertation
to be made available to the students and staff of the University.

\bigskip
\leftline{Signed}

\medskip
\leftline{Date}

\newpage

\section*{Proforma}

{\large
\begin{tabular}{ll}
Candidate Number:   & \bf 2320D \\
Project Title:      & \bf Implementing Parallelism in Lisp for REDUCE \\
Examination:        & \bf Computer Science Tripos -- Part II, 2019  \\
Word Count:         & \bf 12256          \\
Final code line count: & \bf 3686        \\
Project Originator: & Dr. Arthur C. Norman                    \\
Supervisor:         & Dr. Arthur C. Norman                    \\
\end{tabular}
}

\subsection*{Original Aims of the Project}

The goal was to implement a multi-threaded version of a Lisp system. It should
allow the user to create threads and offer the primitives necessary to
write parallel programs. The original language is able to build and
run a complete computer algebra system, which the parallel version should
be able to replicate fully. Additionally, a number of tests and algorithms need to
be implemented to evaluate correctness and the potential benefits of parallelism.

\subsection*{Work Completed}

All of the original goals were achieved. The multi-threaded Lisp
is showing predictable speed-ups when executing tasks in parallel,
while preserving full backwards compatibility with the original language.
It is also able to run the full REDUCE Algebra System, with a number of experiments
showing the potential benefits of parallelism in computer algebra.

\subsection*{Special Difficulties}
There were no special difficulties encountered.

\newpage

\section*{Acknowledgements}
I would like to thank Dr Arthur Norman for all the invaluable help
and support he offered during the making of this project.

\setcounter{tocdepth}{2}
\tableofcontents

\newpage

% Chapters
%=============================================================

\pagestyle{headings}

\setcounter{page}{0}
\pagenumbering{arabic}

\chapter{Introduction}
\label{ch:introduction}
\input{chapters/introduction}

\chapter{Preparation}
\label{ch:preparation}
\input{chapters/preparation}

\chapter{Implementation}
\label{ch:implementation}
\input{chapters/implementation}

\chapter{Evaluation}
\label{ch:evaluation}
\input{chapters/evaluation}

\chapter{Conclusion}
\label{ch:conclusion}
\input{chapters/conclusion}

% Bibliography and Appendix
%=============================================================

\addcontentsline{toc}{chapter}{Bibliography}
\printbibliography

\appendix

% \chapter{Glossary}
% \input{appendix/glossary}

\chapter{Source Code Listings}
\label{ch:appendix-code}
\input{appendix/code}

\chapter{Gröbner basis background}
\label{ch:grobner}
\input{appendix/grobner.tex}

\chapter{Project Proposal}
\input{appendix/proposal}

\end{document}
